---
title: "EDPS 845 Assignment 1"
author: "Brooke Milstead"
---

## Exercises

### Chapter 1


1. Show that you can access the home page for the HTML R help menu, as well as help files for individual functions.

```{r}
#home page for the HTML R help menu
help.start()
```

```{r}
#example help for a function
help("mean")
#OR
?mean
```



2. Demonstrate correct and incorrect naming of R objects.

```{r, eval = FALSE}
#Incorrect variable name examples
incorrect-ex<- 
incorrect_ex <-
incorrect-ex = 
```

```{r, eval = FALSE}
#Correct variable name example
correct.ex <- 
```

```{r, eval = FALSE}
#Incorrect function name examples
subtract_mean <-
subtract-mean<- 
subtractmean = 
```
 
```{r, eval = FALSE}
#Correct function name example
SubtractMean <-
```


3. How do you list all available objects in the R workspace? And how do you remove individual objects?
```{r}
#list available objects
ls() 
#OR
objects()
```

```{r, eval = FALSE}
#remove objects
rm(data.ch1)
```


### Chapter 2


1. Create and index a vector using a logical, integer, negative integer, and character vector.

```{r}
#Create a vector
data.ch2<- c(9, 38, 5, 41, 33, 8, 21, 1, 49, 18)
```

```{r}
#Index a vector- Logical

logical.vector <- data.ch2 > 25
class(logical.vector)
logical.vector

index.True <- data.ch2[data.ch2 > 25]
index.False <- data.ch2[data.ch2 <= 25]

index.True
index.False
```

```{r}
#Index a vector- Integer
integer.vector <- as.integer(data.ch2)
index.int <- data.ch2[3:7]

index.int
```

```{r}
#Index a vector- Negative Integer

index.neg.int <- data.ch2[-(3:7)]

index.neg.int
```

```{r}
#Index a vector- Character 

names(data.ch2) <- c("a", "b", "c", "d", "e", "f", "g", "h", "i", "j")
vowels <- data.ch2[c("a", "e", "i")]

vowels
```

2. Create a simple data frame containing a numeric variable, character variable, and factor with at least 3 levels. 
Use the data frame to demonstrate how R codes and handles missing data. 

```{r}
#numeric variable
num.var <- c(24, 49, NA, 29, 14, 38, 41, 0/0, 10, 33)
```

```{r}
#character variable 
char.var <- c((paste(c("M"), 1:5, sep = "")), 
  (paste(c("F"), 1:5, sep = "")))
class(char.var)
```

```{r}
#factor with at least 3 levels
factor.var <- factor(c("dem", "dem", "rep", "ind", "dem", "rep", "rep", "ind", "dem", "rep" ))

class(factor.var)
levels(factor.var)
```

```{r}
#Create a data frame
Q2.data.frame <- data.frame(num.var, char.var, factor.var)

Q2.data.frame

#NA is a marker for missing values or values not availabe
#NaN represents numerical computations that result in an invalid number
```




### Chapter 3

1. Demonstrate how to check and change the modes and attributes of R objects.

```{r}
#check mode
mode(num.var) 

#change mode example
c.num.var <- as.character(num.var)
mode(c.num.var) 
```

```{r}
#check attributes
attributes(Q2.data.frame)

#changing class to list example
Q2.list <- unclass(Q2.data.frame)
class(Q2.list)
```



### Chapter 4

1. Complete the exercise on page 17, where you calculate 95% confidence intervals for state mean incomes.


```{r}
#Set up for exercise
state <- c("tas", "sa", "qld", "nsw", "nsw", "nt", "wa", "wa",
"qld", "vic", "nsw", "vic", "qld", "qld", "sa", "tas",
"sa", "nt", "wa", "vic", "qld", "nsw", "nsw", "wa",
"sa", "act", "nsw", "vic", "vic", "act")

statef <- factor(state)

incomes <- c(60, 49, 40, 61, 64, 60, 59, 54, 62, 69, 70, 42, 56,
61, 61, 61, 58, 51, 48, 65, 49, 49, 41, 48, 52, 46,
59, 46, 58, 43)
```

```{r}
#means, standardeviation, standard error
incmeans <- tapply(incomes, statef, mean)
stderr <- function(x) sqrt(var(x)/length(x))
incster <- tapply(incomes, statef, stderr)
ns <- table(statef)
```

```{r}
#table summerizing calulations including confidence intervals
cis <- data.frame(mean = incmeans, stderr = incster,
  state = names(incmeans),
  lower = qt(.025, ns - 1),
  upper = qt(.025, ns - 1, lower.tail = FALSE))
cis
```



2. In what order are the levels of ordered factors stored in?


```{r}
#example of how levels are normally stored unless specified otherwise- Alphabetical
levels(statef)
```



### Chapter 5

1. Create an array with dimensions `c(3, 6, 2)` and use it to demonstrate array indexing.
```{r}
#Creating an array
data.51<-sample(1:100, 36)
dim(data.51) <- c(3, 6, 2)
data.51

#indexing examples
data.51[,,1]
data.51[2,,]
data.51[,5,]
data.51[2,5,1]
```

2. Create a two dimensional array of 100 integers sampled with replacement from the vector 1:20, and use an index array to set the diagonal values to 0.

```{r}
#creating array
data.52 <- array(1:20, dim=c(10,10))
data.52

#index array
i <- array(c(1:10), dim = c(10,2))

#exract elements & replace with 0
data.52[i] <- 0

data.52

```

3. Construct the same array using the `array()` function and using a vector with the `dim()` set. Prove they are the same.

```{r}
#Create values for array
data.53s <- sample(1:20, 100, replace = TRUE)

#Use the array function & dim vector
data.53 <- array(data.53s, dim = c(10, 10))
data.53
```

4. Use your array from 5.1, and another with the same dimensions, to demonstrate how arithmetic works with arrays.

```{r}
#new array
data.new <-sample(1:50, 36)
dim(data.new) <- c(3, 6, 2)
data.new

#addition
add.ex <- data.51 + data.new
add.ex

#multiplication
mult.ex <- data.51 * data.new
mult.ex  

#cross products
cp.ex <- crossprod(data.51, data.new)

#outer product
op.ex<- data.51 %o% data.new
```

5. Show how to clear the dimensions of an array.

```{r}
dim(data.new) <- NULL
```

### Chapter 6

1. Demonstrate the different ways of subsetting a list.

```{r}
#create list
lst.61 <- list(state = state, income = incomes)

#subseting component with numbers
lst.61[[1]]

#subseting component and first entry with numbers
lst.61[[1]][1]

#subsetting using name
lst.61$state
#OR
lst.61[["state"]]

#subsetting using name, first entry
lst.61$state[1]

```

2. What is the difference between giving multiple lists to `list()` versus `c()`?

```{r}
#Using list(), you do not have to have the objects class as list to begin with
list.ex2 <- list(state = state, income = incomes, vect = data.ch2)

#Using c(), the objects must be of class list first in order to have separate components.  Otherwise the objects entered will be one component 
list.A <- list(state = state)
list.B <- list(income = incomes)
list.C <- list(vect = data.ch2)

list.ex2c <- c(list.A, list.B, list.C)
list.ex2c
```

3. How do data frames in R relate to lists?

```{r}
#data frame example
data.63 <- data.frame(state = statef, income = incomes)

#a data frame is a list with some restrictions
#components must be vectors, factors, numeric matrices, lists, or data frames
#must have the same length.  matrix structures must have same row size

```

4. What is the purpose of `attach()`?

```{r}
#attach
attach(data.63)

#attach() places the data base in the R search path allowing objects in the database to be accessed using their names alone instead of the components of the data frame
```

### Chapter 7

1. Demonstrate the different approaches to reading in data from an external file.

```{r}
#using read.table
data.71RT <- read.table("DataEX.data.txt", sep = "", header = FALSE)

#using read.csv
read.csv("DataEX.data.csv", header = FALSE)

#using read.delim
read.delim("DataEX.data.txt", sep = "\t", header = FALSE)


```

```{r, eval = FALSE}
#using data from a package
library(equate)
PISA
```


### Chapter 8

1. Demonstrate the use of distribution functions to simulate data and calculate probabilities and quantiles.

```{r}
#simulate data- normal distribution
norm.ex <- rnorm(50, 100, 15)
summary(norm.ex)

#simulate data- exponential distribution
exp.ex <- rexp(50)
summary(exp.ex)

#simulate data- uniform distribution
unif.ex <- runif(50, 1, 20)
summary(unif.ex)
```


```{r}
#p-value, 2-tailed, normal distribution
2*pnorm(2.1)

#upper 10% for an F(3, 12) distribution
qf(0.10, 3, 12, lower.tail = FALSE)

```



2. Simulate data from a t distribution and check their normality.

```{r}
#simulate data
sim.tdata <- rt(50, df=5)
summary(sim.tdata)

#check normality- shapiro
shapiro.test(sim.tdata)


```

```{r}

#check normality- normal q-q plot
plot.qq <- qqnorm(sim.tdata)
plot.qq.l <- qqline(sim.tdata)

#check normality- shapiro
shapiro.test(sim.tdata)

```


### Chapter 9

help(if)

1. What is the difference between `if() else` and `ifelse()`?

```{r}
#if() else evaluates the first element in the expression, if it does not adhear to the logical argument, then the condition after else is evident

#example if() else
age <- 23
if(age <= 19) {print("teens")}else {print("young adults")}

#ifelse() is vectorized. A comparison is made using the logical argument and values are returned for the true and false elements
data.9.1 <- c(3:-3)
ex.9.1 <- ifelse(data.9.1 >= 0, sqrt(data.9.1), NA)
ex.9.1
```

2. Demonstrate the use of a `for()` loop.

```{r}
#create data for example
data.92 <- as.integer(runif(20, 10, 30))

#for loop example
for (i in c(13, 23, 14, 28, 21, 17)) {
if(i <= 19) {print("teens")}else {print("young adults")}}

```
